---
title: "Model update"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(Hmisc)
library(rms)
library(glmnet)
library(tidyr)
library(dplyr)
library(purrr)
library(ROCit)
```

This notebook updates iteratively updates the base prognostic model applying recalibration, by adding additional lab variables in blocks of 2 days

Some ideas to consider further:
* If Elastic Net variable selection is applied on data w/out missing data,
  generate frames by censoring a given percentage of data (e.g.: 10%) and apply MI
* Try Bayesian Variable Selection:
https://github.com/BasisResearch/millipede
https://cran.r-project.org/web/packages/BayesVarSel/
https://cran.r-project.org/web/packages/BayesVarSel/BayesVarSel.pdf

```{r message=FALSE, cache=F}
source('./mdl-evaluation.R')
source('./mdl-selection.R')

num_imp <-100       # Number of multiple imputations
num_boot <- 200     # Number of bootstrap samples
random_seed <- 151  # Setting random seed for reproducibility

min_age <- 18       #  Minimum age for inclusion in analysis

max_cor <- 0.75     # Maximum correlation between variables to be included in model
```

```{r}
# Load the data, remove children, remove patients with less than 1 day of stay, and transform Ct values into Z-scores

data <- read.csv('data.csv')
data <- data %>% filter(age >= min_age)
data <- data %>% filter(days > 0)

data$ct12 <- (data$ct12 - mean(data$ct12, na.rm=T))/sd(data$ct12, na.rm=T)
data$ct34 <- (data$ct34 - mean(data$ct34, na.rm=T))/sd(data$ct34, na.rm=T)
data$ct56 <- (data$ct56 - mean(data$ct56, na.rm=T))/sd(data$ct56, na.rm=T)
data$ct78 <- (data$ct78 - mean(data$ct78, na.rm=T))/sd(data$ct78, na.rm=T)

data
```

```{r}
# Construct the base prognostic model that only uses age and Ct in the first 48 hours

base_model <- function(df) {
  # Calculate the linear term in the logistic function of the model
  # p(x) = 1 / (1+exp(-lp(x))
  lp = with(df, 
            0.86215 - 
              0.05343*age + 
              3*10^-5*max(age - 5.0, 0)^3 - 
              6*10^-5*max(age - 30.0, 0)^3 + 
              3*10^-5*max(age - 58.0, 0)^3 - 
              0.8694*ct12)
  return(lp)
}
```

```{r}
# Add the logit contribution from the base model to the dataframe

vars0 <- c('out', 'age', 'ct12')
all_data0 <- data %>% select(all_of(vars0)) # This dataframe includes missing entries

lp0 <- unlist(map(1:nrow(all_data0), function(x) base_model(all_data0[x,])))
data['lp0'] <- lp0
data
```


```{r}
# Plot calibration and ROC plots of the base model

mdl_folder0 <- "model-0"
dir.create(mdl_folder0)

data0 <- all_data0[complete.cases(all_data0),]

pred0 <- unlist(map(1:nrow(data0), function(x) base_model(data0[x,]))) # prediction in log-odds (probabilities on logit scale)
prob0 <- 1 / (1 + exp(-pred0))                                         # prediction as probability response

# Generate the data frame with the data needed for the calibration and ROC plots
plot_data0 <- data.frame(out = data0$out, logodds = pred0, prob = prob0)

cal_plot(base_model, plot_data0, "prob", model_label = '', dir = mdl_folder0, fn = "cal_plot.pdf")
cal_c0 <- cal_coeff(plot_data0, "prob", mdl_folder0, "cal_plot_coefficients.txt")
print(cal_c0)

roc_plot(data0$out, prob0, dir = mdl_folder0, fn = "roc_plot.pdf")
```

```{r}
plot_roc_perf(plot_data0$out, plot_data0$prob)
```

```{r}
conf_matrix(plot_data0$out, plot_data0$prob, 0.5, mdl_folder0, "conf_matrix.txt")
```

==============================================================================================
UPDATED MODEL FOR DAYS 1-2

```{r}
# Apply the Elastic Net variable selecion method on all the lab variables available for days 1-2

sat_imp_formula12 <- "~out + lp0 + alb12 + alt12 + ast12 + bun12 + ck12 + cre12 + crp12 + k12"

vars12 <- c('out', 'lp0', 'alb12', 'alt12', 'ast12', 'bun12', 'ck12', 'cre12', 'crp12', 'k12')
data12 <- data[complete.cases(data[, c('lp0')]),]
data12 <- data12 %>% filter(days > 0)
data12

sout12 <- selectVariables(random_seed, num_imp, sat_imp_formula12, data12, TRUE)

varf12 <- data.frame(unlist(sout12$names), unlist(sout12$counts))
names(varf12) <- c("Variable", "Freq")
varf12 <- varf12 %>% filter(Variable != "lp0" & Variable != "(Intercept)")
varf12
```

```{r}
# Apply threshold
day12_sel_thres <- 0.5
sel12 <- varf12 %>% filter(Freq > day12_sel_thres)
sel12
```

```{r}
cov12 <- cor(data12[,sel12$Variable], use='pairwise') > 0.75
cov12
```

```{r}
library(BayesVarSel)


formula <- as.formula(sat_imp_formula12)
imp_data <- aregImpute(formula, data=data12, n.impute=1)
for (i in 1:1) {
    df <- getImpute(impute=imp_data, df=data12, im=i)
    m <- Bvs(formula = out ~ lp0 + alb12 + alt12 + ast12 + bun12 + ck12 + cre12 + crp12 + k12, data = df)
    print(m)
}
```


```{r}
# Train the model for days 1-2 on the selected variables
mdl_vars12 <- append(sel12$Variable, 'lp0')
# mdl_vars12 <- mdl_vars12[! mdl_vars12 %in% c()] # Remove highly correlated variables
mdl_vars12

mdl_imp_formula12 <- "~out + lp0 + alb12 + alt12 + ast12 + bun12 + ck12 + cre12 + crp12 + k12"
mdl_lgr_formula12 <- paste('out', paste(mdl_vars12, collapse=" + "), sep=" ~ ")
mdl_lgr_formula12

mdl_folder12 <- "model-days12"
dir.create(mdl_folder12)

model12 <- generateModel(random_seed, num_imp, num_boot, 3, mdl_imp_formula12, mdl_lgr_formula12, data12, mdl_vars12, mdl_folder12)

# Add the logit contribution from the model for days 1-2 to the dataframe
data['lp12'] <- predict(model12, data)
```

```{r}
# Generate the calibration and ROC plots for the models for days 1-2

vars <- append(mdl_vars12, 'out')
df <- data %>% select(all_of(vars))
full_data12 <- df[complete.cases(df),]
full_data12

pred12 <- predict(model12, full_data12) # prediction in log-odds (probabilities on logit scale)
probs12 <- 1 / (1 + exp(-pred12))  # prediction as probability response

# Generate the data frame with the data needed for the calibration and ROC plots
plot_data12 <- data.frame(out = full_data12$out, logodds = pred12, prob = probs12)

cal_plot(model12, plot_data12, "prob", model_label = '', dir = mdl_folder12, fn = "cal_plot.pdf")
cal_c12 <- cal_coeff(plot_data12, "prob", mdl_folder12, "cal_plot_coefficients.txt")
print(cal_c12)

roc_plot(plot_data12$out, plot_data12$prob, dir = mdl_folder12, fn = "roc_plot.pdf")
```

```{r}
plot_roc_perf(plot_data12$out, plot_data12$prob)
```

```{r}
conf_matrix(plot_data12$out, plot_data12$prob, 0.5, mdl_folder12, "conf_matrix.txt")
```

==============================================================================================
UPDATED MODEL FOR DAYS 3-4

```{r}
# Apply the Elastic Net variable selecion method on all the lab variables available for days 3-4

sat_imp_formula34 <- "~out + lp12 + alb34 + alt34 + ast34 + bun34 + ck34 + cre34 + crp34 + k34"

vars34 <- c('out', 'lp12', 'alb34', 'alt34', 'ast34', 'bun34', 'ck34', 'cre34', 'crp34', 'k34')
data34 <- data[complete.cases(data[, c('lp12')]),]
data34 <- data34 %>% filter(days > 2)
data34

sout34 <- selectVariables(random_seed, num_imp, sat_imp_formula34, data34, TRUE)

varf34 <- data.frame(unlist(sout34$names), unlist(sout34$counts))
names(varf34) <- c("Variable", "Freq")
varf34 <- varf34 %>% filter(Variable != "lp12" & Variable != "(Intercept)")
varf34
```

```{r}
# Apply threshold
day34_sel_thres <- 0.5
sel34 <- varf34 %>% filter(Freq > day34_sel_thres)
sel34
```

```{r}
cov34 <- cor(data34[,sel34$Variable], use='pairwise') > 0.75
cov34
```

```{r}
library(BayesVarSel)

formula <- as.formula(sat_imp_formula34)
imp_data <- aregImpute(formula, data=data34, n.impute=1)
for (i in 1:1) {
    df <- getImpute(impute=imp_data, df=data34, im=i)
    m <- Bvs(formula = out ~ lp12 + alb34 + alt34 + ast34 + bun34 + ck34 + cre34 + crp34 + k34, data = df)
    print(m)
}
```


```{r}
# Train the model for days 3-4 on the selected variables
mdl_vars34 <- append(sel34$Variable, 'lp12')
mdl_vars34 <- mdl_vars34[! mdl_vars34 %in% c('cre34')] # Remove highly correlated variables
mdl_vars34

mdl_imp_formula34 <- "~out + lp12 + alb34 + alt34 + ast34 + bun34 + ck34 + cre34 + crp34 + k34"
mdl_lgr_formula34 <- paste('out', paste(mdl_vars34, collapse=" + "), sep=" ~ ")
mdl_lgr_formula34

mdl_folder34 <- "model-days34"
dir.create(mdl_folder34)

model34 <- generateModel(random_seed, num_imp, num_boot, 3, mdl_imp_formula34, mdl_lgr_formula34, data34, mdl_vars34, mdl_folder34)

# Add the logit contribution from the model for days 3-4 to the dataframe
data['lp34'] <- predict(model34, data)
```

```{r}
# Generate the calibration and ROC plots for the models for days 3-4

vars <- append(mdl_vars34, 'out')
df <- data %>% select(all_of(vars))
full_data34 <- df[complete.cases(df),]
full_data34

pred34 <- predict(model34, full_data34) # prediction in log-odds (probabilities on logit scale)
probs34 <- 1 / (1 + exp(-pred34))  # prediction as probability response

# Generate the data frame with the data needed for the calibration and ROC plots
plot_data34 <- data.frame(out = full_data34$out, logodds = pred34, prob = probs34)

cal_plot(model34, plot_data34, "prob", model_label = '', dir = mdl_folder34, fn = "cal_plot.pdf")
cal_c34 <- cal_coeff(plot_data34, "prob", mdl_folder34, "cal_plot_coefficients.txt")
print(cal_c34)

roc_plot(plot_data34$out, plot_data34$prob, dir = mdl_folder34, fn = "roc_plot.pdf")
```

```{r}
plot_roc_perf(plot_data34$out, plot_data34$prob)
```

```{r}
conf_matrix(plot_data34$out, plot_data34$prob, 0.4, mdl_folder34, "conf_matrix.txt")
```

==============================================================================================
UPDATED MODEL FOR DAYS 5-6

```{r}
# Apply the Elastic Net variable selecion method on all the lab variables available for days 5-6

sat_imp_formula56 <- "~out + lp34 + alb56 + alt56 + ast56 + bun56 + ck56 + cre56 + crp56 + k56"

vars56 <- c('out', 'lp34', 'alb56', 'alt56', 'ast56', 'bun56', 'ck56', 'cre56', 'crp56', 'k56')
data56 <- data[complete.cases(data[, c('lp34')]),]
data56 <- data56 %>% filter(days > 4)
data56

sout56 <- selectVariables(random_seed, num_imp, sat_imp_formula56, data56, TRUE)

varf56 <- data.frame(unlist(sout56$names), unlist(sout56$counts))
names(varf56) <- c("Variable", "Freq")
varf56 <- varf56 %>% filter(Variable != "lp34" & Variable != "(Intercept)")
varf56
```

```{r}
# Apply threshold
day56_sel_thres <- 0.5
sel56 <- varf56 %>% filter(Freq > day56_sel_thres)
sel56
```

```{r}
cov56 <- cor(data56[,sel56$Variable], use='pairwise') > 0.75
cov56
```

```{r}
library(BayesVarSel)

formula <- as.formula(sat_imp_formula56)
imp_data <- aregImpute(formula, data=data56, n.impute=1)
for (i in 1:1) {
    df <- getImpute(impute=imp_data, df=data56, im=i)
    m <- Bvs(formula = out ~ lp34 + alb56 + alt56 + ast56 + bun56 + ck56 + cre56 + crp56 + k56, data = df)
    print(m)
}

```

```{r}
# Train the model for days 5-6 on the selected variables
mdl_vars56 <- append(sel56$Variable, 'lp34')
#mdl_vars56 <- mdl_vars56[! mdl_vars56 %in% c()] # Remove highly correlated variables
mdl_vars56

mdl_imp_formula56 <- "~out + lp34 + alb56 + alt56 + ast56 + bun56 + ck56 + cre56 + crp56 + k56"
mdl_lgr_formula56 <- paste('out', paste(mdl_vars56, collapse=" + "), sep=" ~ ")
mdl_lgr_formula56

mdl_folder56 <- "model-days56"
dir.create(mdl_folder56)

# setting number of knots to zero based on this discussion:
# https://stats.stackexchange.com/questions/117383/impute-missing-values-using-aregimpute
model56 <- generateModel(random_seed, num_imp, num_boot, 0, mdl_imp_formula56, mdl_lgr_formula56, data56, mdl_vars56, mdl_folder56)

# Add the logit contribution from the model for days 5-6 to the dataframe
data['lp56'] <- predict(model56, data)
```

```{r}
# Generate the calibration and ROC plots for the models for days 5-6

vars <- append(mdl_vars56, 'out')
df <- data %>% select(all_of(vars))
full_data56 <- df[complete.cases(df),]
full_data56

pred56 <- predict(model56, full_data56) # prediction in log-odds (probabilities on logit scale)
probs56 <- 1 / (1 + exp(-pred56))  # prediction as probability response

# Generate the data frame with the data needed for the calibration and ROC plots
plot_data56 <- data.frame(out = full_data56$out, logodds = pred56, prob = probs56)

cal_plot(model56, plot_data56, "prob", model_label = '', dir = mdl_folder56, fn = "cal_plot.pdf")
cal_c56 <- cal_coeff(plot_data56, "prob", mdl_folder56, "cal_plot_coefficients.txt")
print(cal_c56)

roc_plot(plot_data56$out, plot_data56$prob, dir = mdl_folder56, fn = "roc_plot.pdf")
```

```{r}
plot_roc_perf(plot_data56$out, plot_data56$prob)
```

```{r}
conf_matrix(plot_data56$out, plot_data56$prob, 0.6, mdl_folder56, "conf_matrix.txt")
```

==============================================================================================
UPDATED MODEL FOR DAYS 7-8

DOES NOT WORK BECAUSE THE DATA IS HIGHLY CLASS-IMBALANCED (out of 39 cases, 37 survived are 2 died, leading to variable selection/model fitting to fail)

```{r}
# Apply the Elastic Net variable selecion method on all the lab variables available for days 7-8

sat_imp_formula78 <- "~out + lp56 + alb78 + alt78 + ast78 + bun78 + ck78 + cre78 + crp78 + k78"

vars78 <- c('out', 'lp56', 'alb78', 'alt78', 'ast78', 'bun78', 'ck78', 'cre78', 'crp78', 'k78')
data78 <- data[complete.cases(data[, c('lp56')]),]
data78 <- data78 %>% filter(days > 6)
data78

sout78 <- selectVariables(random_seed, num_imp, sat_imp_formula78, data78, TRUE)

varf78 <- data.frame(unlist(sout78$names), unlist(sout78$counts))
names(varf78) <- c("Variable", "Freq")
varf78 <- varf78 %>% filter(Variable != "lp56" & Variable != "(Intercept)")
varf78
```

```{r}
# Apply threshold
day78_sel_thres <- 0.5
sel78 <- varf78 %>% filter(Freq > day78_sel_thres)
sel78
```

```{r}
cov78 <- cor(data78[,sel78$Variable], use='pairwise') > 0.75
cov78
```

```{r}
# Train the model for days 7-8 on the selected variables
mdl_vars78 <- append(sel78$Variable, 'lp56')
#mdl_vars78 <- mdl_vars78[! mdl_vars78 %in% c()] # Remove highly correlated variables
mdl_vars78

mdl_imp_formula78 <- "~out + lp56 + alb78 + alt78 + ast78 + bun78 + ck78 + cre78 + crp78 + k78"
mdl_lgr_formula78 <- paste('out', paste(mdl_vars78, collapse=" + "), sep=" ~ ")
mdl_lgr_formula78

mdl_folder78 <- "model-days78"
dir.create(mdl_folder78)

# setting number of knots to zero based on this discussion:
# https://stats.stackexchange.com/questions/117383/impute-missing-values-using-aregimpute
model78 <- generateModel(random_seed, num_imp, num_boot, 0, mdl_imp_formula78, mdl_lgr_formula78, data78, mdl_vars78, mdl_folder78)

# Add the logit contribution from the model for days 7-8 to the dataframe
data['lp78'] <- predict(model78, data)
```

```{r}
# Generate the calibration and ROC plots for the models for days 7-8

vars <- append(mdl_vars78, 'out')
df <- data %>% select(all_of(vars))
full_data78 <- df[complete.cases(df),]
full_data78

pred78 <- predict(model78, full_data78) # prediction in log-odds (probabilities on logit scale)
probs78 <- 1 / (1 + exp(-pred78))  # prediction as probability response

# Generate the data frame with the data needed for the calibration and ROC plots
plot_data78 <- data.frame(out = full_data78$out, logodds = pred78, prob = probs78)

cal_plot(model78, plot_data78, "original", "prob", model_label = '', dir = mdl_folder78, fn = "cal_plot.pdf")
roc_plot(plot_data78$out, plot_data78$prob, dir = mdl_folder78, fn = "roc_plot.pdf")
```